Bayesian Hierarchical Model Predicts Dim Prospects for Trump's Victory in the 2024 U.S. Presidential Election

### 2024-11-04 paper/paper.qmd
prose problem fixed discussion part more edited sketch for is_national and pct uploaded

In this paper, we focus on the outcome of the 2024 presidential election. Using a dataset from FiveThirtyEight containing extensive polling results, we examine the influence of various factors on candidate support rates by accounting for differences in polling methods, state-level variation, and temporal changes. It is evident that polling methods often fail to fully capture the complexity of public opinion, and factors that introduce bias may intervene from the very design of polling methods. The chaos and complexity of the real world are also reflected in a large-scale political event like the 2024 presidential election, which involves a vast number of participants, wide-ranging groups, and massive investments and interests. Therefore, we employ a hierarchical Bayesian model as a predictive tool in our analysis. This model integrates polling data from FiveThirtyEight covering various states and polling organizations, while also considering the effects of regional and methodological influences. By incorporating variables such as days since Biden’s withdrawal, polling methodology, and whether the poll is national or state-level, the model provides a dynamic view of Trump’s support rate across time and place. With an R² of 0.641 and an RMSE of 2.31, the model effectively captures these patterns while maintaining predictive accuracy, indicating that it can provide meaningful insights into regional and national public opinion trends. Finally, we simulate data for Election Day on November 5 to provide a reliable basis for predicting the election outcome.

Our analysis reveals that differences in Trump’s support rate across regions and polling methods are particularly pronounced. For example, support rates in traditionally conservative states like Texas differ significantly from those in traditionally liberal states like California, while trends in swing states such as Pennsylvania and Michigan are more unstable, with the potential to disrupt the election balance. This underscores the critical impact of geographic and methodological factors on public opinion, aligning well with the historical tendency of swing states to determine the election outcome. Additionally, our results suggest that polling methodologies introduce systematic differences in support rate estimation, with slight variations between results from online panels and phone surveys. In the predictive results, support for Trump at both the state and national levels hovers around 45% to 50%. Comparing this to the support rates of past U.S. presidents, we predict that Trump is unlikely to win in this election.

@@

### Outcome Variables
In this analysis of Donald Trump’s support rate in the 2024 U.S. presidential election, the outcome variable is Trump’s support rate (i.e., pct) because it is directly related to the election results. The primary research goal of this analysis is to predict Trump’s chances of winning in the 2024 election, so the support rate is the main metric we aim to model and forecast. The support rate here represents the proportion of respondents in a specific poll who support Trump. We seek to understand how the support rate is influenced by different factors, such as polling methodology, time, or geographical variables. By modeling the fluctuations in Trump’s support rate, we can infer his electoral performance, and assess the impact of various factors on his popularity. Understanding these dynamics provides insights into Trump’s standing relative to his opponents and supports our overall election prediction.

@@

```{r, message=FALSE, fig.width= 4, fig.height= 3}
#| label: fig-pct3
#| fig-cap: National support rate and state level support rate for Donald Trump 
#| echo: false

ggplot(analysis_data, aes(x = factor(is_national, levels = c(0, 1), labels = c("State Polls", "National Polls")), 
                 y = pct)) +
  geom_boxplot() +
  labs(x = "Poll Type", y = "Trump Support Rate (%)", title = "Support Rate by Poll Type (National vs. State)") +
  theme_minimal()
```

@@

The linear component of the model is defined as follows:

$$
\mu_i = \alpha + \beta_1 \cdot \text{methodology}_i+ \beta_2 \cdot \mathrm{days\_since\_Biden\_Withdrawal}_i \\
       + \beta_3 \cdot \mathrm{is\_national}_i + u_{\text{state}[i]} + (v_{\text{pollster}[i]} + \gamma_{\text{pollster}[i]} \cdot \mathrm{days\_since\_Biden\_Withdrawal}_i)
$$

@@

Through the `is_national` variable, we distinguish between national and state-level polls, this variable enables the model to capture support rate differences arising from geographic and demographic differences. We expect that national polls, which typically have larger and more representative samples, are better suited to capturing overall trends, while state-level polls may reflect finer, localized biases and specific regional issues because national polls often encompass a broader and more balanced voter base, they provide a more stable forecast of overall voter trends. In contrast, due to geographic and demographic variations, state-level polls often exhibit greater fluctuations in error than national polls. These fluctuations stem from regional issues and diverse demographic characteristics, which may not align fully with national trends [see @kennedy2018disentangling].

@@

# Results {#sec-results}
## Model Output
Our results are summarized in a table which is available in @tbl-model. This summary reveals the effects of different polling methods and organizations on our estimates. For instance, there is a small difference in estimates between data collected via online panels and phone surveys. The model adjusts for these differences by using a hierarchical structure, which accounts for inconsistencies caused by methodological and organizational biases. Specifically, variance parameters for different methods indicate the different impacts these methods have on sampling coverage and voter behavior, helping the model balance these differences.

@@

## Prediction Output
The swig states information is from @270toWin. @fig-results shows that states with similar demographic characteristics (such as Pennsylvania and Michigan) exhibit comparable fluctuations in support rates. It also indicates that the model predicts Trump’s national support rate on Election Day, November 5, 2024, to center around 45% to 50%, with moderate variation. This suggests that despite the influence of national events and polling method differences, the national support rate remains relatively stable, reflecting a steady stance among most voters. We forecast that Trump’s national average support rate will be below 50%.

At the state level, the chart reveals variations in support rates among states, with swing states other than Arizona and North Carolina tending to show average support rates below 50%. Swing states like Pennsylvania and Michigan display significant volatility, as indicated by wider interquartile ranges and multiple outliers in the box plot. This volatility suggests that voter preferences in these states are less stable, and support rates in competitive states are more susceptible to minor events or policy shifts. Therefore, we predict Trump’s support rate in swing states other than Arizona and North Carolina will remain below 50%.

As key factors in the election, this forecast data, combined with Harris’s lead, suggests that Trump’s chance of winning 2024 presidential election is slim.

@@

## Winner Prediction {#sec-first}
Based on the support rate predictions for Trump at both the national and state levels displayed in @fig-results in @sec-results, it appears challenging for Trump to emerge victorious in the upcoming U.S. presidential election on November 5, 2024, against his primary opponent, Harris. The first chart shows the national support rate prediction for Trump, centering around 45% to 50%. Historically, any candidate needs to reach or exceed the 50% threshold to secure a decisive victory. The current predictions reveal some level of uncertainty, although minor shifts in voter sentiment might occur leading up to Election Day, leaving Trump with some possibility of winning.

The @fig-results also illustrates the support rates at the state level, revealing significant differences between red and blue states. For instance, Trump’s predicted support rate is higher in traditionally conservative states such as Texas, while in strong Democratic states like California, his support rate is lower. In swing states like Pennsylvania and Michigan, the support rate hovers around 50%, aligning with historical trends where these states tend to have close contests.

To draw final conclusions from these predictions, it is essential to consider historical polling data and past election results. In previous U.S. presidential elections, shifts in sentiment in swing states have played a important role in determining the outcome. For example, the 2016 election highlighted the importance of swing states like Pennsylvania, Michigan, and Wisconsin, where polls underestimated Trump’s support. A study by the Pew Research Center noted that last-minute sentiment shifts and polling errors led to an unexpected advantage for Trump [@pew2021pollinglimitations]. Studies by *The Economist* emphasize the critical impact of swing states on election outcomes, also noting that national polls do not always reflect the dynamics of the Electoral College [see @economist2020electionforecast]. The Electoral College system can result in a candidate winning the popular vote but losing the election, or vice versa. This system has affected U.S. elections multiple times, demonstrating the importance of state-level results relative to national polling trends.

@@

### Defects of Existing Data
Firstly, a major challenge for the model lies in estimating a large number of parameters with limited polling data. This scarcity of data can make the model highly sensitive to choices in priors and interstate covariance structures. For example, relying on historical patterns from states with limited data may fail to reflect the latest shifts in voter sentiment. This issue is also seen across different polling organizations and survey methods. As @gelman2020regression noted, the model can fluctuate between overconfidence and unreasonable uncertainty. To address this, we could consider incorporating alternative data sources, such as social media sentiment or historical population voting trends, to supplement polling data in underrepresented areas. However, these sources carry their own issues, as they may lack reliability and rigor.

Then, in handling systematic bias introduced by different polling methods, the model accounts for variance by assigning specific variance parameters for each method (e.g., phone interviews vs. online panels). Each method may systematically overestimate or underestimate support for a particular candidate. For instance, online panels may attract more politically engaged individuals, leading to biases compared to random phone interviews. To reduce these biases, the model includes variance adjustments specific to each method. Further improvements could involve re-weighting polls based on historical accuracy or introducing dynamic adjustments for newer polling methods. However, as pointed out in @blumenthal2010polls, historical data often include responses to unique events that are not repeatable, making it difficult to assess the accuracy of historical information as a reference. Similarly, the model uses a random walk prior to accommodate fluctuations in public support over time, yet it may struggle to capture abrupt changes driven by political events, such as a debate or controversial statement. The inherent randomness of world events makes it challenging to fully address this problem.

@@

### The Influence of Reality
#### Bias and Transparency
The influence of polling agencies’ biases and methodological transparency on election predictions has long been a research focus. Every polling agency's choices—whether intentional or unintentional—can introduce biases into results. In highly polarized political environments, factors like question wording, sampling techniques, and demographic weighting can amplify potential biases. For example, research from the [Pew Research](https://www.pewresearch.org/short-reads/2024/08/28/key-things-to-know-about-us-election-polling-in-2024/) Center and [Stanford University](https://news.stanford.edu/stories/2020/10/9-things-know-election-polling-data) shows that differences in methods, such as educational weighting or nonresponse adjustments, can lead to significant polling biases, especially for politically sensitive figures like Donald Trump [see @pew2024electionpolling and @stanford2020electionpolling].

@@

#### Financial Constraints
Financial limitations are a major barrier to achieving ideal polling coverage, particularly when trying to gather representative polling data across all U.S. states. Polling each state comprehensively is costly, especially in low-population states or states deemed "safe" for one candidate. In these cases, data scarcity means the model largely relies on data from more populous or swing states, which may not fully represent the diversity of the nationwide electorate, introducing potential biases.

@@

Visualizations often struggle to fully capture the complex structure of our model. The model includes data from 50 U.S. states, multiple polling organizations, and various survey methodologies. If we attempted to display every detail within a chart, the visualizations would become overwhelmingly complex, making it difficult for readers to extract key insights. Conversely, if we limit the display to a subset, such as only highlighting key states or selected polling methods, this approach risks losing the completeness of information. These simplified visualizations fail to convey all sources of variation within the model, potentially leading readers to overlook significant factors and misunderstand the impact of omitted influences. Additionally, this partial representation can foster a feedback loop where oversimplified summaries influence potential voters' perceptions, impacting the trends in unpredictable ways.

@@

@fig-post5 shows the posterior distributions of certain parameters in our model, including the `(Intercept)`, a methodology variable `methodologyIVR/Online Panel/Text-to-Web`, a time variable `days_since_Biden_Withdrawal`, and the error term `sigma`. From the shape and location of the posterior distributions, we can see the degree to which the observed data supports the estimates of these parameters and the range of uncertainty around them. The `(Intercept)` distribution is centered around 30% to 55%, suggesting that the model estimates a 45% baseline support rate, which aligns with our expectations for Trump’s support rate. The methodology variable shows a relatively wide distribution, indicating that the data does not provide strong directional support for its influence, suggesting that it may have a minimal or uncertain impact on the support rate. Also, the time variable has a posterior distribution close to zero, indicating that the number of days since Biden’s withdrawal does not significantly impact the support rate. The error term `sigma` is at a low level, suggesting that the model views the data as having low volatility and being relatively concentrated. Thus, we can conclude that the model effectively captures the significant impact of the baseline support rate but but due to data uncertainty, it struggles to reveal the influence of factors like methodology and time. This phenomenon may reflect inconsistencies in the data across different poll methods or limitations in the the sample.

>>>>

In this paper, we focus on the outcome of the 2024 presidential election. Using a dataset from FiveThirtyEight containing extensive polling results, we examine the influence of various factors on candidate support rates by accounting for differences in polling methods, state-level variation, and temporal changes. It is evident that polling methods often fail to fully capture the complexity of public opinion, and factors that introduce bias may intervene from the very design of polling methods. The chaos and complexity of the real world are also reflected in a large-scale political event like the 2024 presidential election, which involves a vast number of participants, wide-ranging groups, and massive investments and interests. Therefore, we employ a hierarchical Bayesian model as a predictive tool in our analysis. This model integrates polling data from FiveThirtyEight covering various states and polling organizations, while also considering the effects of regional and methodological influences. By incorporating variables such as days since Biden’s withdrawal, polling methodology, and whether the poll is national or state-level, the model provides a dynamic view of Trump’s support rate across time and place. With an R² of 0.641 and an RMSE of 2.31, the model effectively captures these patterns while maintaining predictive accuracy, indicating that it can provide a reliable prediction for regional and national public opinion trends. Finally, we simulate data for Election Day on November 5 to provide a reliable basis for predicting the election outcome.

Our analysis shows that differences in Trump’s support rate across regions and polling methods are particularly pronounced. For example, support rates in traditionally conservative states like Texas differ significantly from those in traditionally liberal states like California, while trends in swing states such as Pennsylvania and Michigan are more unstable, with the potential to disrupt the election balance. This underscores the impact of geographic and methodological factors on public opinion, aligning well with the historical tendency of swing states to determine the election outcome. Additionally, our results suggest that polling methodologies introduce systematic differences in support rate estimation, with slight variations between results from online panels and phone surveys. In the predictive results, support for Trump at both the state and national levels hovers around 45% to 50%. Comparing this to the support rates of past U.S. presidents, we predict that Trump is unlikely to win in this election.

@@

### Outcome Variables
In this analysis of Donald Trump’s support rate in the 2024 U.S. presidential election, the outcome variable is Trump’s support rate (i.e., pct) because it is directly related to the election results. The primary research goal of this analysis is to predict Trump’s chances of winning in the 2024 election, so the support rate is the main metric we aim to model and forecast. The support rate here represents the proportion of respondents in a specific poll who support Trump. We seek to understand how the support rate is influenced by different factors, such as polling methodology, time, or geographical variables. By modeling the fluctuations in Trump’s support rate, we can infer his electoral performance, and assess the impact of various factors on his popularity. Understanding these dynamics supports our overall election prediction.

@@

```{r, message=FALSE, fig.width= 5, fig.height= 3}
#| label: fig-pct3
#| fig-cap: National support rate and state level support rate for Donald Trump 
#| echo: false

ggplot(analysis_data, aes(x = factor(is_national, levels = c(0, 1), labels = c("State Polls", "National Polls")), 
                 y = pct)) +
  geom_boxplot() +
  labs(x = "Poll Type", y = "Trump Support Rate (%)", title = "Support Rate by Poll Type (National vs. State)") +
  theme_minimal() + theme(
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 5),
    plot.title = element_text(size = 7, face = "bold", color = "#333333"),
    axis.title = element_text(size = 6, color = "#333333"),
  )
```

@@

The linear component of the model is defined as follows (here $\mathrm{days\_since\_Biden\_Withdrawal}$ is shortened to DSBW):

$$
\mu_i = \alpha + \beta_1 \cdot \text{methodology}_i+ \beta_2 \cdot \mathrm{DSBW}_i \\
\\
       + \beta_3 \cdot \mathrm{is\_national}_i + u_{\text{state}[i]} + (v_{\text{pollster}[i]} + \gamma_{\text{pollster}[i]} \cdot \mathrm{DSBW}_i) \\
$$

@@

Through the `is_national` variable, we distinguish between national and state-level polls, this variable enables the model to capture support rate differences arising from geographic and demographic differences. We expect that national polls, which typically have larger and more representative samples, are better suited to capturing overall trends, while state-level polls may reflect finer, localized biases and specific regional issues because national polls often include a broader and more balanced voter base, they provide a more stable forecast of overall voter trends. In contrast, due to geographic and demographic variations, state-level polls often exhibit greater fluctuations in error than national polls. These fluctuations stem from regional issues and diverse demographic characteristics, which may not align fully with national trends [see @kennedy2018disentangling].

@@

# Results {#sec-results}
## Model Output
Our results are summarized in a table which is available in @tbl-model. This summary shows the effects of different polling methods and organizations on our estimates. For instance, there is a small difference in estimates between data collected via online panels and phone surveys. The model adjusts for these differences by using a hierarchical structure, which accounts for inconsistencies caused by methodological and organizational biases. Specifically, variance parameters for different methods indicate the different impacts these methods have on sampling coverage and voter behavior, helping the model balance these differences.

@@

## Prediction Output
The swig states information is from @270toWin. @fig-results shows that states with similar demographic characteristics (such as Pennsylvania and Michigan) exhibit comparable fluctuations in support rates. It also indicates that the model predicts Trump’s national support rate on Election Day, November 5, 2024, to center around 45% to 50%, with moderate variation. This suggests that despite the influence of national events and polling method differences, the national support rate remains relatively stable, reflecting a steady stance among most voters. We forecast that Trump’s national average support rate will be below 50%. At the state level, @fig-results shows variations in support rates among swing states, with swing states other than Arizona and North Carolina tending to show average support rates below 50%. Swing states like Pennsylvania and Michigan display significant volatility, as indicated by wider interquartile ranges and multiple outliers in the box plot. This volatility suggests that voter preferences in these states are less stable, and support rates in competitive states are more susceptible to minor events or policy shifts.

@@

## Winner Prediction {#sec-first}
Based on the support rate predictions for Trump at both the national and state levels displayed in @fig-results in @sec-results, it appears challenging for Trump to emerge victorious in the upcoming U.S. presidential election on November 5, 2024, against his primary opponent, Harris. @fig-results shows the national support rate prediction for Trump, centering around 45% to 50%. Historically, any candidate needs to reach or exceed the 50% threshold to secure a decisive victory. The current predictions show some level of uncertainty, although minor shifts in voter sentiment might occur leading up to Election Day, leaving Trump with some possibility of winning.

The @fig-results also illustrates the support rates at the state level. In swing states like Pennsylvania and Michigan, the support rate mostly under 50%, aligning with historical trends where these states tend to have close contests but have big influence on the final election results. We predict Trump’s support rate in swing states other than Arizona and North Carolina will remain below 50%. Therefore, as important factors in the election, this forecast data, combined with Harris’s lead, suggests that Trump’s chance of winning 2024 presidential election is slim.

We draw final decision based on previous U.S. presidential elections, where shifts in sentiment in swing states have played a important role in determining the outcome. For example, the 2016 election highlighted the importance of swing states like Pennsylvania, Michigan, and Wisconsin, where polls underestimated Trump’s support. A study by the Pew Research Center noted that last-minute sentiment shifts and polling errors led to an unexpected advantage for Trump [@pew2021pollinglimitations]. Studies by *The Economist* emphasize the important impact of swing states on election outcomes, also noting that national polls do not always reflect the dynamics of the Electoral College [see @economist2020electionforecast]. The Electoral College system can result in a candidate winning the popular vote but losing the election, or vice versa. This system has affected U.S. elections multiple times, demonstrating the importance of state-level results relative to national polling trends.

@@

### Defects of Existing Data
Firstly, a major challenge for the model lies in estimating a large number of parameters with limited polling data. This scarcity of data can make the model highly sensitive to choices in priors and interstate covariance structures. For example, relying on historical patterns from states with limited data may fail to reflect the latest shifts in voter sentiment. To address this, we could consider incorporating alternative data sources in future, such as social media sentiment or historical population voting trends, to supplement polling data in underrepresented areas. However, these sources carry their own issues, as they may lack reliability and rigor.

Similar problem is also seen across different polling organizations and survey methods. In handling systematic bias introduced by different polling methods, the model accounts for variance by assigning specific variance parameters for each method (e.g., phone interviews vs. online panels). Each method may systematically overestimate or underestimate support for a particular candidate. For instance, online panels may attract more politically engaged individuals, leading to biases compared to random phone interviews. To reduce these biases, the model includes variance adjustments specific to each method. Further improvements could involve re-weighting polls based on historical accuracy or introducing dynamic adjustments for newer polling methods. However, as pointed out in @blumenthal2010polls, historical data often include responses to unique events that are not repeatable, making it difficult to assess the accuracy of historical information as a reference. Similarly, the model uses a random walk prior to accommodate fluctuations in public support over time, yet it may struggle to capture abrupt changes driven by political events, such as a debate or controversial statement. The inherent randomness of world events makes it challenging to fully address this problem.

@@

### The Influence of Reality
#### Bias and Transparency
The influence of polling agencies’ biases and methodological transparency on election predictions has long been a research focus. Every polling agency's choices—whether intentional or unintentional—can introduce biases into results. In highly polarized political environments, factors like question wording, sampling techniques, and demographic weighting can amplify potential biases. For example, research from the [Pew Research](https://www.pewresearch.org/short-reads/2024/08/28/key-things-to-know-about-us-election-polling-in-2024/) Center and [Stanford University](https://news.stanford.edu/stories/2020/10/9-things-know-election-polling-data) shows that differences in methods, such as educational weighting or nonresponse adjustments, can lead to significant polling biases [see @pew2024electionpolling and @stanford2020electionpolling].

@@

#### Financial Constraints
Financial limitations are a major barrier to achieving ideal polling coverage, particularly when trying to gather representative polling data across all U.S. states. Polling each state is costly, especially in low-population states or states deemed "safe" for one candidate. In these cases, data scarcity means the model largely relies on data from more populous or swing states, which may not fully represent the diversity of the nationwide electorate, introducing potential biases.

@@

Visualizations often struggle to fully capture the complex structure of our model. The model includes data from 50 U.S. states, multiple polling organizations, and various survey methodologies. If we attempted to display every detail within a chart, the visualizations would become overwhelmingly complex, making it difficult for readers to read or understand. Conversely, if we limit the display to a subset, such as only highlighting key states or selected polling methods, this approach risks losing the completeness of information. These simplified visualizations fail to convey all sources of variation within the model, potentially leading readers to overlook significant factors and misunderstand the impact of omitted influences. Additionally, this partial representation can foster a feedback loop where oversimplified summaries influence potential voters' perceptions, impacting the trends in unpredictable ways.

@@

@fig-post5 shows the posterior distributions of certain parameters in our model, including the `(Intercept)`, a methodology variable `methodologyIVR/Online Panel/Text-to-Web`, a time variable `days_since_Biden_Withdrawal`, and the error term `sigma`. From the shape and location of the posterior distributions, we can see the degree to which the observed data supports the estimates of these parameters and the range of uncertainty around them. The `(Intercept)` distribution is centered around 30% to 55%, suggesting that the model estimates a 45% baseline support rate, which aligns with our expectations for Trump’s support rate. The methodology variable shows a relatively wide distribution, indicating that the data does not provide strong directional support for its influence, suggesting that it may have a minimal or uncertain impact on the support rate. Also, the time variable has a posterior distribution close to zero, indicating that the number of days since Biden’s withdrawal does not significantly impact the support rate. The error term `sigma` is at a low level, suggesting that the model views the data as having low volatility and being relatively concentrated. Thus, we can conclude that the model effectively captures the significant impact of the baseline support rate but but due to data uncertainty, it struggles to catch the influence of factors like methodology and time. This phenomenon may reflect inconsistencies in the data across different poll methods or limitations in the the sample.
